---
title: "Book recommender"
output: html_notebook
author: "Ondrej Havlicek"
---

This is an attempt at a book recommendation system.

More specifically, the assignment is: I am a reader, who likes Cloud Atlas and Lord of the Rings. Recommend me what to read next.

The simplest approach might use just some attributes of the books, like genre, year, average rating at goodreads.. (= content-based?)

A better approach is to use other users' ratings to estimate what a given person might like based on similarity between users or between items (= collaborative filtering).

First we need a database of book ratings: downloaded from [IIF Freiburg](http://www2.informatik.uni-freiburg.de/~cziegler/BX/). 

## Prepare ratigns data

Libraries, functions
```{r Libraries}
library(tidyverse)

getCosine <- function(x,y) 
{
  return( sum(x*y) / (sqrt(sum(x*x)) * sqrt(sum(y*y))) )
}

```

Settings, constants, etc..
Cloud Atlas is the book by David Mitchell. There is also a book "The Cloud Atlas" by Liam Callanan which we don't want.
Lord of the Rings is actually a series of three books, include all of them and the original term as well, because there might be a book/product which is called this way, e.g. a collection of all the books.
```{r}
datafolder <- file.path(getwd(),"data")

myfavs <- c("Cloud Atlas", "The Lord of the Rings", "The Fellowship of the Ring", "The Two Towers", "The Return of the King")
myfavsshort <- c("CA", "LOTR", "FOTR", "TTT", "ROTK")
```

## Prepare data
Read data
```{r}
Users <- read_delim(file.path(datafolder, "BX-Users.csv"), ";", quote = "\"", na="NULL", col_types = "ici")

Books <- read_delim(file.path(datafolder, "BX-Books.csv"), ";", quote = "\"", na="NULL", col_types = "cccicccc")

#Ratings <- read_delim(file.path(datafolder, "BX-Book-Ratings.csv"), ";", quote = "\"", na="NULL", col_types = "ici")
Ratings <- read_delim(file.path(datafolder, "BX-Book-Ratings.csv"), ";", quote = "\"", na="NULL", col_types = "ici")  
```
There were some parsing problems, ignore for now..


Modify data
```{r}
Users <- Users %>% separate(col=Location, into=c("Loc1", "Loc2", "Loc3"), sep=",")

Books <- Books %>% select(ISBN, `Book-Title`, `Book-Author`, `Year-Of-Publication`)

Ratings <- Ratings %>% filter(`Book-Rating` > 0) # Remove implicit ratings (coded as 0) now: those mean that the person has purchased/read the book, but we know nothing about how much he/she liked it so it cannot be used in the same way as the explicit ratings, but perhaps later we could include it..
```
Location format apparently not always consistent, but we won|t be using this information for now anyway


Select only the books we care about. The same book may have multiple name variations and ISBNs, so the surest way is to do it semi-manually.
```{r}
findbooktitles <- adist(Books$`Book-Title`, myfavs)  # "fuzzy" string matching
findbooktitles <- as_tibble(findbooktitles) 
colnames(findbooktitles) <- myfavsshort

MyBooks <- bind_cols(Books, findbooktitles) 
#glimpse(MyBooks)

MyBooks %>% arrange(CA) %>% head(50)  # Find the correct Cloud Atlas - cannot find it
#MyBooks %>% filter(tolower(`Book-Author`) == "david mitchell")  # encoding problems apparently, too lazy to solve now..
MyBooks %>% filter(`Book-Author` %in% c("david mitchell", "David Mitchell", "DAVID MITCHELL", "Mitchell, David"))  # ah, it is "Cloud Atlas : A Novel"

caISBN <- c("0375507256")
myISBNs <- caISBN

MyBooks %>% arrange(LOTR) %>% head(50)  # the top few records are what we want, all have distance <= 2
lotrISBNs <- MyBooks %>% filter(LOTR <= 2) %>% .$ISBN

MyBooks %>% arrange(FOTR) %>% head(50)  # distance = 0
lotrISBNs <- c(lotrISBNs, MyBooks %>% filter(FOTR == 0) %>% .$ISBN)

MyBooks %>% arrange(TTT) %>% head(50)  # distance = 0
lotrISBNs <- c(lotrISBNs, MyBooks %>% filter(TTT == 0) %>% .$ISBN)

MyBooks %>% arrange(ROTK) %>% head(50)  # distance = 0
lotrISBNs <- c(lotrISBNs, MyBooks %>% filter(ROTK == 0) %>% .$ISBN)

myISBNs <- c(caISBN,lotrISBNs)

MyBooks <- MyBooks %>% filter(ISBN %in% myISBNs) %>% select(-(CA:ROTK))
```
That's quite a few ISBNs for LOTR, we may need to care about giving equal weight to Cloud Atlas, but will see..

Find all the ratings given to these books
```{r}
MyBooksRatings <- left_join(MyBooks, Ratings, by="ISBN")

```


Does the DB allow multiple ratings of the same book by the same user?
```{r}
#MyBooksRatings %>% filter(ISBN %in% myISBNs) %>% arrange(`User-ID`)
MyBooksRatings %>% filter(ISBN %in% myISBNs) %>% group_by(`User-ID`, ISBN) %>% summarize(count = n()) %>% ungroup() %>% summarize(maxcount = max(count))
```
Probably not.

How frequent are which ratings?
```{r}
MyBooksRatings %>% count(`Book-Rating`)  
with(MyBooksRatings, table(ISBN, `Book-Rating`)) 
myISBNsWithRatings <- MyBooksRatings %>% filter(!is.na(`Book-Rating`)) %>% group_by(ISBN) %>% summarize(count = length(`Book-Rating`)) %>% filter(count > 0) %>% .$ISBN  # count only those books that have some non-NA ratings (the latter part is redundant because of how MyBooksRatings were created, but for sure..)
```
Most ratings are in the range 8-10
3 of the books were never rated


## Create a user-book rating matrix
```{r}
#RM <- Ratings %>% spread(`User-ID`, `Book-Rating`) #Error: invalid input '349800924ß' in 'utf8towcs' - some encoding error??
#Ratings <- Ratings %>% mutate(ISBNa = iconv(enc2utf8(ISBN), sub = "byte")) #convert encoding
#RM <- Ratings %>% select(-ISBN) %>% spread(`User-ID`, `Book-Rating`) #Error: cannot allocate vector of size 267.1 Gb

``` 
Cannot fit the whole matrix into memory..

Limit only to those users who have rated at least one of the selected books.
And to books which have a rating by at least one of those users (so that the matrix does not have any empty rows)
```{r}
usersWithRatings <- MyBooksRatings %>% select(`User-ID`) %>% arrange() %>% unique() %>% .$`User-ID`
#length(usersWithRatings)  # only 60 users have rated at least one of the CA and LOTR books..
MyRatings <- Ratings %>% filter(`User-ID` %in% usersWithRatings)
# transform Ratings from long to wide format = rating matrix
RM <- MyRatings %>% spread(`User-ID`, `Book-Rating`)  # the usual format seems to be with users in rows and items in columns, but I did it in the opposite way
RM <- RM %>% filter_at(vars(-ISBN), any_vars(!is.na(.)))  # select only rows with at least one rating
rm <- RM %>% select(-ISBN) %>% as.matrix()  # matrix should be faster to work with than tibble
#dim(rm) #7751 60
dimnames(rm) <- list(RM$ISBN, RM %>% select(-ISBN) %>% colnames())  # rows, columns

rm.na <- rm  # wasting memory, but for now save the rating matrix including the NA values into a separate variable
rm[is.na(rm)] <- 5.5  # for now simply replace all missing values with 5.5 = middle value, probably better than zero, that would bias the missing ratings as "low"; or we could use the mean rating of a user, or a different approach altogether

#we could also normalize the ratings, but often one user only gives ratings of 10 and normalization would make them 0..
```

Now the real computation needs to happen:-)

## Compute item-based similarity

Item similarit is more meaningful than user similarity and we don't know the active user's ratings anyway
Which similarity measure to use? Pearson correlation is a good one
Cosine similarity is another option
It has to deal with all the missing values..

```{r}
# Explore how to do it
# Compute similarities not all with all, but only our books of interest with the other books
# RM.ca <- RM %>% filter(ISBN == caISBN)
# RM.lotr <- RM %>% filter(ISBN %in% lotrISBNs)
# RM.other <- RM %>% filter(!(ISBN %in% myISBNs))  
# 
# #cor(as.numeric(RM.ca[1,-1]), as.numeric(RM.other[1,-1]), use="na.or.complete")
# 
# cors <- numeric(nrow(RM.other))
# 
# for (i in 1:nrow(RM.other)) {
#   cors[i] <- cor(as.numeric(RM.ca[1,-1]), as.numeric(RM.other[i,-1]), use="na.or.complete")
# }

# Or try it with the matrix
# Try to compute similarity of CA with all books
#rm[caISBN,]  # CA was rated only once??
sims <- numeric(nrow(rm))

for (i in 1:nrow(rm)) {
  #sims[i] <- cor(rm[1,], rm[i,], use="na.or.complete")
  sims[i] <- getCosine(rm[caISBN,], rm[i,]) #seems to be more sensitive to differences in ratings than the correlation
}

RM$cosineCA <- sims
top10 <- RM %>% arrange(desc(cosineCA)) %>% head(10)
#Books %>% filter(ISBN %in% top10$ISBN)
top10 %>% select(ISBN, cosineCA) %>% left_join(Books, by="ISBN") %>% arrange(desc(cosineCA))
```
So the book most similar book to Cloud Atlas seems to be On the Road by Jack Kerouac, but based apparently on only one rating:-)


Compute a full similarity matrix, to try this more general approach, although we don't need all the similarities
```{r}
#initialize similarity matrix
sm <- matrix(data=NA, nrow = nrow(rm), ncol = nrow(rm), dimnames = list(RM$ISBN, RM$ISBN) )
#now the main heavy lifting happens, could be probably optimized somehow, now just compute one half of the matrix, because it is symmetric
for (i in 1:nrow(rm)) {
  for (j in i:nrow(rm)) {
    sm[i,j] <- getCosine(rm[i,], rm[j,])
  }
}
sm[is.na(sm)] <- t(sm)[is.na(sm)] #I googled a way to fill the other half of the matrix:)
# sm has 791 MB??
```


## Make the recommendations
```{r}
#select rows with input books
myIndexes <- RM$ISBN %in% myISBNsWithRatings
mySM <- sm[,myIndexes]  # list of similarities of our input books to all other books
mySM.ISBNs <- dimnames(mySM)[[1]]
mySM <- as_tibble(mySM)

# We don't know the users' ratings for CA and LOTR (and thus can't use them as weights), we just know he/she likes them
# for now just take the mean of the similarities, i.e. giving the same weight to CA and all the (many) LOTR books, although this now biases the results in favor of LOTR similarity
#mySM %>% summarize_all(funs(MeanSimil = mean(.)))
mySM$MeanSimil <- rowMeans(mySM)
#mySM <- mySM %>% mutate(MeanSimil2 = rowMeans(.[,1:16]) )
#mySM$MeanSimil <- rowMeans(mySM[,1:length(myISBNsWithRatings)])
mySM$ISBN <- mySM.ISBNs  # label the rows explicitly
#Final answer?:)
Recommendations <- mySM %>% select(ISBN, MeanSimil) %>% arrange(desc(MeanSimil)) %>% left_join(Books, by="ISBN")
Recommendations %>% head(50)
#Makes no sense.. how are these books similar to LOTR?
Recommendations %>% filter(ISBN %in% myISBNsWithRatings)  # relatively lower mutual similarity ratings, weird

```

The recommendations we got are somewhat strange. It might be due to the 15 different ISBNs for the LOTR books, it might help to unify them and their ratings, manually or per book title (&author) instead of per ISBN.
There are apparently many ways how to do recommendation systems. Some of the tasks seem to be quite standard so not surprisingly there are packages for them:


## Try to use a package rather than doing it manually
```{r}
library(recommenderlab)
# Prepare rating matrix
rm.rl <- t(rm.na)  # these methods like to have users as rows and items as columns
rm.rl <- as(rm.rl,"realRatingMatrix")  # convert the matrix into their specific format
# Normalize it, but problem may be that there may not be enough ratings per user to properly scale the ratings..
#rm.rl <- normalize(rm.rl, method="center")

# Create an active user vector, assume ratings of 10 for books he/she likes
activeUser <- matrix(NA, nrow=1, ncol=ncol(rm.rl), dimnames = list("active", dimnames(rm.rl)[[2]]) )  #as(rm.rl[1,], "matrix")
ISBNs <- dimnames(activeUser)[[2]]  # all ISBNS in the matrix/model
existingISBNs <- myISBNsWithRatings[myISBNsWithRatings %in% ISBNs]  # not all of the books we care about were rated by someone = are not in the matrix/model
activeUser[,existingISBNs] <- 10  # Assign CA and all the LOTR books rating of 10
activeUser <- as(activeUser, "realRatingMatrix")
# Add the new user to the rating matrix
rma <- rbind(as(rm.rl, "matrix"), as(activeUser, "matrix"))
rma <- as(rma, "realRatingMatrix")

# Create Item-based model, takes a while
ibModel <- Recommender(rma,method="IBCF", 
                       param=list(method="Cosine", k=50))  # can we NOT do normalization? the method doesnt seem to be well documented

```

Get recommendations
```{r}
#ibPrediction <- predict(ibModel, activeUser, n=10)
ibPrediction <- predict(ibModel, rma["active",], type="ratings")
as(ibPrediction, "list") #we get nothing, why?
ibPrediction <- predict(ibModel, rma[1,], type="ratings") #some random user
as(ibPrediction, "list") #we get something
recommendedBooks <- predict(ibModel, rma[1,], type="topN", n=10)
Books %>% filter(ISBN %in% as(recommendedBooks, "list")[[1]] )


```
We don't get any recommendation, possibly because all the ratings were 10 and they got normalized to 0, we could try a different method, try to remove the normalization, or add some fake ratings of 1 for some random/fake book not rated by anyone else.
The possibilities are wide..






